A Formal Model of Checked C

### 2022-07-15 sandp23/formal.tex
update.

@@

% LEO: This has an overfull line...?
\begin{figure}[t]
{\small
{\captionsetup[lstlisting]{margin = 8 mm}
  \begin{lstlisting}[xleftmargin=8 mm]
nt_array_ptr<char> safe_strcat
   (nt_array_ptr<char> dst : count(n),
    nt_array_ptr<char> src : count(0), int n) {
  int x = strlen(dst);
  int y = strlen(src);
  nt_array_ptr<char> c : count(n) =
    dyn_bounds_cast
           <nt_array_ptr<char>>(dst,count(n));
    // sets c == dst with bound n (not x)
  if (x+y < n) {
    for (int i = 0; i < y; ++i)
      *(c+x+i) = *(src+i);
    *(c+x+y) = '\0';
    return dst;
  }
  return null;
}
  \end{lstlisting}
}
}
\caption{Implementation of safe \code{strcat}}
\label{fig:strcat-ex}
\end{figure}

@@

\myparagraph{Checked and Tainted Pointer Operations}
%
The rules for pointer related operations---\textsc{S-DefC},
\textsc{S-DefT}, \textsc{S-AssignArrC}, \textsc{S-AssignArrT},
\textsc{S-DefNull}, \textsc{S-IfNTTC} and \textsc{S-Cast}.
The first five are deference and assignment operations---illustrate how the semantics checks bounds.
Rule \textsc{S-DefNull} transitions attempted null-pointer
dereferences to $\enull$, whereas \textsc{S-DefC} dereferences a \cmode-mode
non-null (single) pointer.
  When $\enull$ is returned by the
computation relation, the evaluation relation halts the entire
evaluation with $\enull$ (using a rule not shown in Fig.~\ref{fig:c-context}); it
does likewise when $\ebounds$ is returned (see \Cref{sec:rem-semantics}).

% \review{- on page 6, section IIIC, paragraph "Pointer Access" mentions that checked pointers cannot be dereferenced in unchecked blocks - this looks funny, shouldn't it be the other way around? The Coq code contains the hypothesis m'=Unchecked -> m Unchecked in various rules of definition well-typed (BoundCheckedC, line 669; rule TyDeref in the code seems closest to the figure's T-DefArr and T-Def in the appendix, although it's a bit concerning that there's no 1-to-1 correspondence of the rules in the code and the paper).
% }
% \liyi{Yeah. It is another typo. It should be the other way around.  }

\textsc{S-AssignArrC} assigns to an array as long as 0 (the point of
dereference) is within the bounds designated by the pointer's annotation
and strictly less than the upper bound. For the assignment rule,
arrays are treated uniformly whether they are null-terminated or not
($\kappa$ can be $\cdot$ or $nt$)---the semantics does not search past
the current position for a null terminator. The program
can widen the bounds as needed, if they currently precede the null
terminator: \textsc{S-DefNTArray}, which dereferences an NT array
pointer, allows an upper bound of $0$, since the program may read, but
not write, the null terminator. A separate rule (not shown) handles
normal arrays.

\myparagraph{Casts}
%
Static casts of a literal $n\!:\!\tau'$ to a type $\tau$ are handled
by \textsc{S-Cast}. In a type-correct program, such casts are
confirmed safe by the type system. To evaluate a cast, the rule
updates the type annotation on $n$. Before doing so, it must
``evaluate'' any variables that occur in $\tau$ according to their
bindings in $\varphi$. For example, if $\tau$ was
$\tarrayptr{0}{x+3}{\tint}{\cmode}$, then $\varphi(\tau)$ would
produce $\tarrayptr{0}{5}{\tint}{\cmode}$ if $\varphi(x) = 2$.

Dynamic casts are accounted for by \textsc{S-DynCast} and
\textsc{S-DynCastBound}. In a type-correct program, such casts are
assumed correct by the type system, and later confirmed by the
semantics. As such, a dynamic cast will cause a $\ebounds$ failure if
the cast-to type is incompatible with the type of the target pointer,
as per the $n_l' > n_l \vee n_h > n_h'$ condition in
\textsc{S-DynCastBound}. An example use of dynamic casts is given on
line~7 in Fig.~\ref{fig:strcat-ex}. The values of \code{x} and
\code{n} might not be known statically, so the type system cannot
confirm that \code{x <= n}; the dynamic cast assumes this inequality
holds, but then checks it at run-time.

@@

\myparagraph{Bounds Widening}
%
Bounds widening occurs when branching on a dereference of a NT array
pointer, or when performing $\estrlentext$. The latter is most useful
when assigned to a local variable so that subsequent code can use the
result, e.g., $e$ in $\elet{x}{\estrlen{y}}{e}$.  Lines 4 and 5 in
Fig.~\ref{fig:strcat-ex} are examples. 
The widened upper bound
precipitated by $\estrlen{y}$ is extended beyond the lifetime of $x$, as long as $y$ is live.
% Even if $x$ returns, if $y$'s upper bound keeps the widened bound that is caused by $\estrlen{y}$.
For example, \code{x}'s scope in line 4 at runtime is
the whole function body in \code{safe_strcat} because the lifetime of the pointer \code{dst} is in the function body.
% Allowing widening to persist
% outside the scope is sound, % but would precipitate the need for
% % fat pointers during compilation, which we want to avoid.
% and we take this approach to give more precise run-time behavior. We
% show in Sec.~\ref{sec:compilation}
This is different from the \checkedc specification, 
which only allows bound widening to happen within the scope of \code{x}, and restoring old bound values once \code{x} dies.
We allow widening to persist outside the scope at run-time as long as
we are within the stack frame, and we show
this does not necessarily require the use of fat pointers in
Sec.~\ref{sec:compilation}.

@@

Rule \textsc{S-IfNTT} performs bounds widening on $x$ when the
dereference $\estar{x}$ is not at the null terminator, but the
pointer's upper bound is 0 (i.e., it's at the end of its known
range). $x$'s upper bound is incremented to 1, and this count persists
as long as $x$ is live.  For example, \code{s}'s increment (lines
5--8) is live until the return of the function in
Fig.~\ref{fig:checkedc-example}; thus, line 11 is valid because
\code{s}'s upper bound is properly extended. %  For this rule and
% \textsc{S-StrWiden}, this widening persists in the current stack
% frame. When $x$ goes out of scope, .

>>>>

\begin{figure}[t]
{\small
  \begin{lstlisting}[xleftmargin=4 mm]
nt_array_ptr<const char>
parse_utf16_hex(nt_array_ptr<const char> s,
                  ptr<uint> result) {
 int x1, x2, x3, x4;
 if (s[0] != 0) { x1 = hex_char_to_int(s[0]);
 if (s[1] != 0) { x2 = hex_char_to_int(s[1]);
 if (s[2] != 0) { x3 = hex_char_to_int(s[2]);
 if (s[3] != 0) { x4 = hex_char_to_int(s[3]);
 if (x1 != -1 && x2 != -1 && x3 != -1 && x4 != -1){
   *result = (uint)((x1<<12)|(x2<<8)|(x3<<4)|x4);
   return s+4;
  ...// several } braces
 } 
 return 0;
}
void parse(nt_array_ptr<const char> s,
            array_ptr<uint> p : count(n), 
            int n) {
 array_ptr<uint> q : bounds(p,p+n) = p;
 while (s && q < p+n) {
   array_ptr<uint> r : count(1) =
     dyn_bounds_cast<array_ptr<uint>>(q,count(1));
   s = parse_utf16_hex(s,r);
   q++;
 }
}
  \end{lstlisting}

% {\captionsetup[lstlisting]{margin = 8 mm}
%   \begin{lstlisting}[xleftmargin=8 mm]
% array_ptr<int> memcpy(int x,
%      array_ptr<int> a : count(x),
%      array_ptr<int> b : count(x)) : count(x) {
%  for(int i = 0; i < x; ++i) {
%     (*a = *b); a++; b++;
%  }
%  return a;
}

% static int idx = 0;
% int buf (int * reg,
%           nt_array_ptr<int> x,
%           nt_array_ptr<int> y) {
%   if (idx < n)
%     unchecked { * x = &(reg + idx); idx++; }
%   else bug();
%   memcpy(n,x,y);
%   return 0;
% }
% }
\caption{Parsing a String of UTF16 Hex Characters in \checkedc}
\label{fig:checkedc-example}
\end{figure}

@@

% LEO: This has an overfull line...?

@@

\myparagraph{Checked and Tainted Pointer Operations}
%
The rules for pointer related operations---\textsc{S-DefC},
\textsc{S-DefT}, \textsc{S-AssignArrC}, \textsc{S-AssignArrT},
\textsc{S-DefNull}, \textsc{S-IfNTTC} and \textsc{S-Cast}.
The first five are deference and assignment operations---illustrate how the semantics checks bounds.
Rule \textsc{S-DefNull} transitions attempted null-pointer
dereferences to $\enull$, whereas \textsc{S-DefC} dereferences a \cmode-mode
non-null (single) pointer.
  When $\enull$ is returned by the
computation relation, the evaluation relation halts the entire
evaluation with $\enull$ (using a rule not shown in Fig.~\ref{fig:c-context}); it
does likewise when $\ebounds$ is returned (see \Cref{sec:rem-semantics}).
\textsc{S-AssignArrC} assigns to an array as long as 0 (the point of
dereference) is within the bounds designated by the pointer's annotation
and strictly less than the upper bound.

\textsc{S-DefC} and \textsc{S-AssignArrT} are the deference and assignment rules for tainted pointers.
Any dynamic heap use of a tainted pointer requires to perform a verification.
Performing such a verification equates to performing a type check for a pointer constant in \Cref{fig:const-type}.
We explain this shortly in \Cref{sec:type-system}.
For now, the verification step, e.g. $\umode;\emptyset;\heap ; \emptyset \vdash\evalue{n_a}{\tau}$ in \textsc{S-DefC},
means we verify that the value $n_a$ has type $\tau$ in the $\umode$ region heap.

Static casts of a literal $n\!:\!\tau'$ to a type $\tau$ are handled
by \textsc{S-Cast}. In a type-correct program, such casts are
confirmed safe by the type system no matter
if the target is a $\tmode$ or $\cmode$ pointer. To evaluate a cast, the rule
updates the type annotation on $n$. Before doing so, it must
``evaluate'' any variables that occur in $\tau$ according to their
bindings in $\varphi$. For example, if $\tau$ was
$\tarrayptr{0}{x+3}{\tint}{\cmode}$, then $\varphi(\tau)$ would
produce $\tarrayptr{0}{5}{\tint}{\cmode}$ if $\varphi(x) = 2$.

Bounds widening occurs when branching on a dereference of a NT array
pointer, or when performing $\estrlentext$, which has been explained in \cite{li22checkedc}. 
Here, we explain Rule \textsc{S-IfNTTC} that 
performs bounds widening on a $\cmode$-mode pointer $x$ when the
dereference $\estar{x}$ is not at the null terminator, but the
pointer's upper bound is 0 (i.e., it's at the end of its known
range). $x$'s upper bound is incremented to 1, and this count persists
as long as $x$ is live.  For example, \code{s}'s increment (lines
5--8) is live until the return of the function in
Fig.~\ref{fig:checkedc-example}; thus, line 11 is valid because
\code{s}'s upper bound is properly extended.

\ignore{
The latter is most useful
when assigned to a local variable so that subsequent code can use the
result, e.g., $e$ in $\elet{x}{\estrlen{y}}{e}$.  Lines 4 and 5 in
Fig.~\ref{fig:strcat-ex} are examples. 
The widened upper bound
precipitated by $\estrlen{y}$ is extended beyond the lifetime of $x$, as long as $y$ is live.
% Even if $x$ returns, if $y$'s upper bound keeps the widened bound that is caused by $\estrlen{y}$.
For example, \code{x}'s scope in line 4 at runtime is
the whole function body in \code{safe_strcat} because the lifetime of the pointer \code{dst} is in the function body.
% Allowing widening to persist
% outside the scope is sound, % but would precipitate the need for
% % fat pointers during compilation, which we want to avoid.
% and we take this approach to give more precise run-time behavior. We
% show in Sec.~\ref{sec:compilation}
This is different from the \checkedc specification, 
which only allows bound widening to happen within the scope of \code{x}, and restoring old bound values once \code{x} dies.
We allow widening to persist outside the scope at run-time as long as
we are within the stack frame, and we show
this does not necessarily require the use of fat pointers in
Sec.~\ref{sec:compilation}.

Rule \textsc{S-IfNTT} performs bounds widening on $x$ when the
dereference $\estar{x}$ is not at the null terminator, but the
pointer's upper bound is 0 (i.e., it's at the end of its known
range). $x$'s upper bound is incremented to 1, and this count persists
as long as $x$ is live.  For example, \code{s}'s increment (lines
5--8) is live until the return of the function in
Fig.~\ref{fig:checkedc-example}; thus, line 11 is valid because
\code{s}'s upper bound is properly extended.
}
% \review{- on page 6, section IIIC, paragraph "Pointer Access" mentions that checked pointers cannot be dereferenced in unchecked blocks - this looks funny, shouldn't it be the other way around? The Coq code contains the hypothesis m'=Unchecked -> m Unchecked in various rules of definition well-typed (BoundCheckedC, line 669; rule TyDeref in the code seems closest to the figure's T-DefArr and T-Def in the appendix, although it's a bit concerning that there's no 1-to-1 correspondence of the rules in the code and the paper).
% }
% \liyi{Yeah. It is another typo. It should be the other way around.  }

\ignore{
\textsc{S-AssignArrC} assigns to an array as long as 0 (the point of
dereference) is within the bounds designated by the pointer's annotation
and strictly less than the upper bound. For the assignment rule,
arrays are treated uniformly whether they are null-terminated or not
($\kappa$ can be $\cdot$ or $nt$)---the semantics does not search past
the current position for a null terminator. The program
can widen the bounds as needed, if they currently precede the null
terminator: \textsc{S-DefNTArray}, which dereferences an NT array
pointer, allows an upper bound of $0$, since the program may read, but
not write, the null terminator. A separate rule (not shown) handles
normal arrays.}

\ignore{
\myparagraph{Casts}
%

Dynamic casts are accounted for by \textsc{S-DynCast} and
\textsc{S-DynCastBound}. In a type-correct program, such casts are
assumed correct by the type system, and later confirmed by the
semantics. As such, a dynamic cast will cause a $\ebounds$ failure if
the cast-to type is incompatible with the type of the target pointer,
as per the $n_l' > n_l \vee n_h > n_h'$ condition in
\textsc{S-DynCastBound}. An example use of dynamic casts is given on
line~7 in Fig.~\ref{fig:strcat-ex}. The values of \code{x} and
\code{n} might not be known statically, so the type system cannot
confirm that \code{x <= n}; the dynamic cast assumes this inequality
holds, but then checks it at run-time.
}

@@

\ignore{
\myparagraph{Bounds Widening}
%
Bounds widening occurs when branching on a dereference of a NT array
pointer, or when performing $\estrlentext$. The latter is most useful
when assigned to a local variable so that subsequent code can use the
result, e.g., $e$ in $\elet{x}{\estrlen{y}}{e}$.  Lines 4 and 5 in
Fig.~\ref{fig:strcat-ex} are examples. 
The widened upper bound
precipitated by $\estrlen{y}$ is extended beyond the lifetime of $x$, as long as $y$ is live.
% Even if $x$ returns, if $y$'s upper bound keeps the widened bound that is caused by $\estrlen{y}$.
For example, \code{x}'s scope in line 4 at runtime is
the whole function body in \code{safe_strcat} because the lifetime of the pointer \code{dst} is in the function body.
% Allowing widening to persist
% outside the scope is sound, % but would precipitate the need for
% % fat pointers during compilation, which we want to avoid.
% and we take this approach to give more precise run-time behavior. We
% show in Sec.~\ref{sec:compilation}
This is different from the \checkedc specification, 
which only allows bound widening to happen within the scope of \code{x}, and restoring old bound values once \code{x} dies.
We allow widening to persist outside the scope at run-time as long as
we are within the stack frame, and we show
this does not necessarily require the use of fat pointers in
Sec.~\ref{sec:compilation}.

@@

Rule \textsc{S-IfNTT} performs bounds widening on $x$ when the
dereference $\estar{x}$ is not at the null terminator, but the
pointer's upper bound is 0 (i.e., it's at the end of its known
range). $x$'s upper bound is incremented to 1, and this count persists
as long as $x$ is live.  For example, \code{s}'s increment (lines
5--8) is live until the return of the function in
Fig.~\ref{fig:checkedc-example}; thus, line 11 is valid because
\code{s}'s upper bound is properly extended.
} %  For this rule and
% \textsc{S-StrWiden}, this widening persists in the current stack
% frame. When $x$ goes out of scope, .
